Processing Script

### 2024-02-20 code/eda-code/eda.qmd
Part 2 Updates
We fixed an error in one of the variables we created, total_larcster. We began our EDA with histograms and scatter plots with numeric variables.

```{r}
#Path to data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed_data","processeddata.rds")
#load data
mydata <- readRDS(data_location)
```

@@

# Data exploration through tables

Showing a bit of code to produce and save a summary table.

```{r}
summary_df = skimr::skim(mydata)
print(summary_df)
# save to file
summarytable_file = here("results", "summarytable.rds")
saveRDS(summary_df, file = summarytable_file)
```

We are saving the results to the `results` folder. Depending on how many tables/figures you have, it might make sense to have separate folders for each. And/or you could have separate folders for exploratory tables/figures and for final tables/figures. Just choose a setup that makes sense for your project and works for you, and provide enough documentation that someone can understand what you are doing.

@@

Histogram plots for the continuous outcomes.

Height first.

```{r}
p1 <- mydata %>% ggplot(aes(x=Height)) + geom_histogram() 
plot(p1)
figure_file = here("results","height_distribution.png")
ggsave(filename = figure_file, plot=p1) 
```

Now weights.

```{r}
p2 <- mydata %>% ggplot(aes(x=Weight)) + geom_histogram() 
plot(p2)
figure_file = here("results","weight_distribution.png")
ggsave(filename = figure_file, plot=p2) 
```

Now height as function of weight.

```{r}
p3 <- mydata %>% ggplot(aes(x=Height, y=Weight)) + geom_point() + geom_smooth(method='lm')
plot(p3)
figure_file = here("results","height_weight.png")
ggsave(filename = figure_file, plot=p3) 
```

@@

>>>>

```{r}
#Path to data
data_location <- here("data","processed-data","processeddata.rds")
#load data
mydata <- readRDS(data_location)
```

@@

# Data exploration through tables
We will begin with creating a summary table of our cleaned variables using the skim()function. This provides information

```{r}
summary_df = skim(mydata)
print(summary_df)
# save to file
summarytable_file = here("results", "tables", "summarytable.rds")
saveRDS(summary_df, file = summarytable_file)
```

@@

We will make histogram plots for the continuous outcomes in our dataset.

Percent of unintended pregnancies is the main outcome of interest, so we will explore this variable first.

```{r}
p1 <- mydata %>% ggplot(aes(x=pct_upreg)) + geom_histogram() 
plot(p1)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_distribution.png")
ggsave(filename = figure_file, plot=p1) 
```

It is important to consider the type of birth control methods associated with percent of unintended pregnancies as well. We will create a histogram for the `pct_upreg_larcster`, `pct_upreg_sarc`, `pct_upreg_trad`, and `pct_upreg_nouse`.

```{r}
p2 <- mydata %>% ggplot(aes(x=pct_upreg_larcster)) + geom_histogram() 
plot(p2)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_larcster_distribution.png")
ggsave(filename = figure_file, plot=p2)

p3 <- mydata %>% ggplot(aes(x=pct_upreg_sarc)) + geom_histogram() 
plot(p3)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_sarc_distribution.png")
ggsave(filename = figure_file, plot=p3)

p4 <- mydata %>% ggplot(aes(x=pct_upreg_trad)) + geom_histogram() 
plot(p4)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_trad_distribution.png")
ggsave(filename = figure_file, plot=p4)

p5 <- mydata %>% ggplot(aes(x=pct_upreg_nouse)) + geom_histogram() 
plot(p5)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_nouse_distribution.png")
ggsave(filename = figure_file, plot=p5) 
```

We will now explore the new 
plots 6-10

We will start exploring how many women use each type of birth control compared to the percent of unintended pregnancies related to that birth control. Plot 11 shows `pct_upreg_larcster` vs `total_larcster` to analyze women using long acting reversible methods and sterilization versus the percentage of unintended pregnancies using those methods.

```{r}
p11 <- mydata %>% ggplot(aes(x=pct_upreg_larcster, y=total_larcster)) + geom_point() + geom_smooth(method='lm')
plot(p11)
figure_file = here("results","height_weight.png")
ggsave(filename = figure_file, plot=p3) 
```

@@

Our next group of plots will be violin plots to explore the usage level of birth associated with percentage of unintended pregnancies use for each birth control method. 
```{r}
p16 <- mydata %>% ggplot(aes(x=larcster_usage, y=pct_upreg_larcster))
```

Our last plots will be mosaic plots to compare two categorical variables: region and usage for each type of birth control. 
```{r}
library(ggmosaic)
p17 <- mydata %>% ggplot(aes(x=region, y=larcster_usage, fill = freq)) + geom_mosaic(aes(x=region, fill = larcster_usage))
```

### 2024-02-20 code/processing-code/processingfile-v1.qmd
Part 2 Updates
We fixed an error in one of the variables we created, total_larcster. We began our EDA with histograms and scatter plots with numeric variables.

```{r}
#| message: false
# select the outcome variable, predictor variables, and covariates  
rawdata2 <- rawdata %>% 
                select(country, region, subregion, pct_upreg, upreg_subreg, pct_upreg_larcster, pct_upreg_sarc, pct_upreg_trad, pct_upreg_nouse, larc_1519, sarc_1519, tradmethods_1519, nomethod_1519, larc_2024, sarc_2024, tradmethods_2024, nomethod_2024, larc_2534,  sarc_2534, tradmethods_2534,  nomethod_2534, larc_3549, sarc_3549, tradmethods_3549, nomethod_3549)

@@

To make the covariate variables more useful, it makes sense to create a new factor variable denoting low, medium, and high usage of each birth control method by country. First, we need to determine the boundaries for each level of usage by exploring how many women use each method across age groups. We need to create a variable showing the total usage of each birth control method to remove the effect of age.  
```{r}
#| message: false
# create new variables showing total usage of each method including all age groups
rawdata2 <- rawdata2 %>% 
              mutate(total_larc = larc_1519 + larc_2024 + larc_2534 + larc_3549,
                     total_sarc = sarc_1519 + sarc_2024 + sarc_2534 + sarc_3549,
                     total_trad = tradmethods_1519 + tradmethods_2024 + tradmethods_2534 + tradmethods_3549,
                     total_none = nomethod_1519 + nomethod_2024 + nomethod_2534 + nomethod_3549)

# explore summary of each total to set boundaries for each level of usage
rawdata2 %>% skim(total_larc, total_sarc, total_trad, total_none)
```
After finding the distribution of number of women using each type of birth control, we can use that information to create a new factor variable denoting low, medium, and high usage for each method. The "low" category will include observations below the 25th percentile. The "medium" category will include observations between the 25th and 75th percentiles. The "high" category will include observations above the 75th percentile. 
```{r}
#| message: false
# create usage variable for long acting reversible methods 
rawdata2 <- rawdata2 %>% 
              mutate(larc_usage = as.factor(case_when(
                                  total_larc < 11.5 ~ "Low",
                                  total_larc >= 11.5 & total_larc < 407.4633 ~ "Medium",
                                  total_larc >= 407.4633 ~ "High")))

# check if new variable was created correctly 
class(rawdata2$larc_usage)

# create usage variable for short acting reversible methods 
rawdata2 <- rawdata2 %>% 
              mutate(sarc_usage = as.factor(case_when(
                                  total_sarc < 114.33113 ~ "Low",
                                  total_sarc >= 114.33113 & total_larc < 1554.7609 ~ "Medium",
                                  total_sarc >= 1554.7609	 ~ "High")))

@@

# create usage variable for traditional methods 
rawdata2 <- rawdata2 %>% 
              mutate(trad_usage = as.factor(case_when(
                                  total_trad < 15.05000	 ~ "Low",
                                  total_trad >= 15.05000 & total_larc < 332.4750 ~ "Medium",
                                  total_trad >= 332.4750	 ~ "High")))

@@

# create usage variable for no methods
rawdata2 <- rawdata2 %>% 
              mutate(none_usage = as.factor(case_when(
                                  total_none < 80.07500	 ~ "Low",
                                  total_none >= 80.07500 & total_larc < 875.2750 ~ "Medium",
                                  total_none >= 875.2750	 ~ "High")))

>>>>

```{r}
#| message: false
# select the outcome variable, predictor variables, and covariates  
rawdata2 <- rawdata %>% 
                select(country, region, subregion, pct_upreg, upreg_subreg, pct_upreg_larcster, pct_upreg_sarc, pct_upreg_trad, pct_upreg_nouse, larc_1519, ster_1519, sarc_1519, tradmethods_1519, nomethod_1519, larc_2024, ster_2024, sarc_2024, tradmethods_2024, nomethod_2024, larc_2534, ster_2534, sarc_2534, tradmethods_2534,  nomethod_2534, larc_3549, ster_3549, sarc_3549, tradmethods_3549, nomethod_3549)

@@

To make the covariate variables more useful, it makes sense to create a new factor variable denoting low, medium, and high usage of each birth control method by country. First, we need to determine the boundaries for each level of usage by exploring how many women use each method across age groups. We need to create a variable showing the total usage of each birth control method to remove the effect of age. It is worth noting here that the long acting reversible and sterilzation methods are combined in the`pct_upreg_larcster` variable, so we need to consider all the women using both of these methods in the total. 
```{r}
#| message: false
# create new variables showing total usage of each method including all age groups
rawdata2 <- rawdata2 %>% 
              mutate(total_larcster = larc_1519 + larc_2024 + larc_2534 + larc_3549 + ster_1519 + ster_2024 + ster_2534 +                          ster_3549,
                     total_sarc = sarc_1519 + sarc_2024 + sarc_2534 + sarc_3549,
                     total_trad = tradmethods_1519 + tradmethods_2024 + tradmethods_2534 + tradmethods_3549,
                     total_none = nomethod_1519 + nomethod_2024 + nomethod_2534 + nomethod_3549)

# explore summary of each total to set boundaries for each level of usage
rawdata2 %>% skim(total_larcster, total_sarc, total_trad, total_none)
```
After finding the distribution of number of women using each type of birth control, we can use that information to create a new factor variable denoting low, medium, and high usage for each method. The "low" category will include observations below the 25th percentile. The "medium" category will include observations between the 25th and 75th percentiles. The "high" category will include observations above the 75th percentile. 
```{r}
#| message: false
# create usage variable for long acting reversible methods 
rawdata2 <- rawdata2 %>% 
              mutate(larcster_usage = as.factor(case_when(
                                  total_larcster < 23.94865 ~ "Low",
                                  total_larcster >= 23.94865 & total_larcster < 1554.7609 ~ "Medium",
                                  total_larcster >= 1554.7609 ~ "High")))

# check if new variable was created correctly 
class(rawdata2$larcster_usage)

# create usage variable for short acting reversible methods 
rawdata2 <- rawdata2 %>% 
              mutate(sarc_usage = as.factor(case_when(
                                  total_sarc < 114.33113 ~ "Low",
                                  total_sarc >= 114.33113 & total_sarc < 1554.7609 ~ "Medium",
                                  total_sarc >= 1554.7609	 ~ "High")))

@@

# create usage variable for traditional methods 
rawdata2 <- rawdata2 %>% 
              mutate(trad_usage = as.factor(case_when(
                                  total_trad < 15.05000	 ~ "Low",
                                  total_trad >= 15.05000 & total_trad < 332.4750 ~ "Medium",
                                  total_trad >= 332.4750	 ~ "High")))

@@

# create usage variable for no methods
rawdata2 <- rawdata2 %>% 
              mutate(none_usage = as.factor(case_when(
                                  total_none < 80.07500	 ~ "Low",
                                  total_none >= 80.07500 & total_none < 875.2750 ~ "Medium",
                                  total_none >= 875.2750	 ~ "High")))
